Climatic warming strengthens a positive feedback\\ between alpine shrubs and fire

### 2016-10-19 manuscript/manuscript.Rnw
More edits to manuscript

\begin{abstract}
Climate change is expected to increase fire activity and woody plant encroachment in arctic and alpine landscapes. However, the extent to which these increases interact to affect changes in structure, function and composition of alpine ecosystems is unknown. Here we use field surveys and experimental manipulations to examine warming and fire effects on recruitment, growth and survival of seedlings of four dominant, evergreen fire-killed alpine shrubs. We find that fire substantially increased shrub seedling establishment (up to \Sexpr{density_ratios(greaus_seedling_density_model, 'Grevillea')}\textendash fold) and that experimental warming doubled tall shrub seedling growth rates and could potentially increase survival. Warming had no effect on shrub recruitment, post-fire infilling of gaps by a dominant alpine grass, or competitive effects of grass on shrub seedling growth and survival. These findings indicate that warming, coupled with more frequent or severe fires, will likely exacerbate increases in tall evergreen shrubs by increasing recruitment, doubling growth rates, and potentially increasing shrub survival. Thus, warming is likely to enhance a positive feedback between woody species abundance and fire activity in alpine and tundra environments.
\end{abstract}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\section*{Introduction}
Accurately forecasting the effects of climatic warming on vegetation dynamics requires an understanding of the mechanisms by which climate and vegetation interact. Most forecasting models include the direct effects of climatic conditions on species distributions, but largely ignore disturbances, the type, frequency and severity of which are also likely to change \citep{Guisan:2005gz,Tucker:2012dr}. Recurrent disturbance is integral to all ecosystems, strongly influencing fundamental demographic processes such as recruitment and mortality, and thus the composition and structure of plant communities and biomes worldwide \citep{Bond:2005eg, Turner:2010wo}. Disturbance regimes are likely to change as a consequence of climate change \citep{Turner:2010wo, Westerling:2011jp,Bradstock:2014jp}. Moreover, the effects of disturbance will interact with climate change, and thus may amplify or diminish how changes to temperature and moisture affect vegetation \citep{Post:2008wz,Camac:2015hja,Enright:2015em}

Feedbacks between climate change, vegetation and fire will be particularly important determinants of the trajectory of ecosystem change in coming decades \citep{Bowman:2009kp}. For example, in addition to the direct effects of warming on fire weather, changing climate may also alter fire regimes by altering vegetation productivity, structure and plant trait composition \citep{Keeley:2012wo,Matthews:2012gia,Bowman:2014dr}. Feedbacks may be positive or negative, and have been documented in tropical savanna \citep{Hoffmann:2003br}, boreal forests \citep{Krawchuk:2011wr}, Mediterranean ecosystems \citep{Keeley:2012wo} and grasslands \citep{Flannigan:2009kl}. Change in both vegetation and fire regimes in alpine and arctic environments are also likely to be subject to feedbacks \citep{Goetz:2007gj,Wookey:2009wo}.

@@

However, most studies in arctic and alpine areas have focused on plant responses in mature, undisturbed vegetation \citep{Briceno:2015iv}. Few climate-change field experiments have included disturbance as a factor \citep[but see:][]{Munier:2010cg,Graae:2010ir,Camac:2015hja}, or examined the influence of climate change on seedlings in post-disturbance environments. Although alpine and tundra vegetation has a strong capacity to regenerate vegetatively \citep{Billings:1968bn} there is mounting evidence that seedling regeneration is important in these ecosystems \citep{Venn:2009bw, Briceno:2015iv}, particularly for wood species \citep{Camac:2013eb,Williams:2014ud}. Although seedlings are highly vulnerable to climate and disturbance, they are the life stage that determines the long-term persistence of a species and its capacity to establish in new areas. However, little is known about how climate change affects seedling demography \citep{Briceno:2015iv}, particularly in post-disturbance conditions in alpine and tundra ecosystems. It is therefore imperative we understand how seedlings respond to both changing climate and disturbance regimes \citep{Walck:2011dm, Briceno:2015iv} if we wish to accurately predict future trajectories of vegetation change.

\section*{Conceptual model}

In alpine and arctic ecosystems, warming experiments and long-term monitoring have documented significant increases in the growth and cover of woody species \citep{Sturm:2001vd,Walker:2006ht,MyersSmith:2011fp}. The frequency and extent of wildfires have also increased in these environments over recent decades, a trend expected to continue \citep{Westerling:2006ve, Flannigan:2009kl, Qiu:2009th, Bradstock:2014jp}. In alpine and tundra ecosystems, current evidence from long-term and experimental studies indicate that shrub recruitment and encroachment is highest in disturbed areas \citep{Williams:1987ws, Batllori:2009fe, Frost:2013tm} and lowest in areas with minimal bare ground cover \citep[e.g. undisturbed alpine tussock grassland;][]{Williams:1987ws}. Current evidence indicates that climatic warming is likely to increase woody species growth rates \citep{Arft:1999iz, Elmendorf:2012ba, MyersSmith:2011fp} and that shrubs are the most flammable vegetation component of these ecosystems \citep{Williams:2006tl, Higuera:2009ga, Fraser:NHT5Kx_W}. Thus, increases in the cover of woody species in alpine and arctic environments is likely to increase the flammability of these ecosystems.

The combination of this evidence indicates a potential positive feedback between warmer temperatures, woody species dynamics, and fire activity in alpine and tundra environments (Fig. \ref{fig1}a). Warmer temperatures are likely to lead to more frequent or severe fire. This is likely to increase woody species recruitment opportunities (i.e. more bare ground) and exacerbate both shrub thickening within shrub communities and shrub expansion into shrub-free communities such as grassland and herbfields \citep{Racine:2004tn,Lantz:2013fh}. If this effect is coupled with increased shrub seedling growth and mortality, highly flammable fuels will accumulate faster and have higher landscape connectivity, ultimately leading to further increases in the likelihood of fire \citep{Higuera:2008dw, Higuera:2009ga,MyersSmith:2011fp}. Thus, warming could strengthen an existing climate-disturbance feedback, that not only has the potential to cause rapid changes in the composition and structure of alpine and arctic vegetation, but also has serious social, biodiversity and carbon sequestration consequences \citep{Mack:2011ew}.

\begin{figure}[!h]
\includegraphics{figures/fig1.pdf}
\caption{Conceptual diagram illustrating positive warming-shrub-fire feedback. (a) Hypothesised positive feedback loop between fire, climatic warming and landscape flammability. Solid lines = known mechansims; Dashed lines = mechansims that may exacerbate or diminish this feedback but which we have a paucity of information on. (b) Updated feedback based on new evidence obtained from experimental and field surveys conducted in this study. Also includes an example of a potential, but unlikely, effect (red dashed line) that could break this feedback. Blue = positive relationships, Red = negative relationships.}
\label{fig1}
\end{figure}

While paleoecological studies have indicated that such a feedback may exist in the arctic \citep{Higuera:2008dw, Higuera:2009ga}, there is a paucity of information on what may strengthen, mitigate or break this feedback. For example, we have little information on whether fire will stimulate shrub seedling recruitment in alpine or tundra ecosystems, nor do we know how rates of seedling growth and survival will be affected under warmer, more exposed, post-fire conditions. Furthermore, we do not know how tussock grasslands (a dominant structural vegetation type in arctic and alpine ecosystems) will respond to warmer post-fire conditions, whether grasses have competitive, facilitative or no effect on shrub seedling vital rates, or whether such effects are altered by warmer post-fire conditions.

Here, we examine these unknowns and their impacts on this proposed feedback between climatic warming, shrubs and fire in Australian alpine vegetation. As with other tundra ecosystems, Australian alpine landscapes consist of a matrix of grass-herb dominanted vegetation types, and shrub-dominanted types \citep{Williams:2014ud}. They have also experienced significant changes in climate. Since 1979, mean growing season temperatures have risen by approximately 0.4\textdegree C and annual precipitation has fallen by 6\% \citep{Wahren:2013ji}, with a consequent decline in snow pack depth \citep{SanchezBayo:2013ci}. These climatic changes have been correlated with a 10 to 20\% increase in shrub cover and a 25\% decline in graminoids cover \citep{Wahren:2013ji}. Much of the Australian Alps has also been burnt by recent (2003 and 2006) wildfires, the frequency and/or severity of which are expected to increase as a consequence of climate change \citep{Bradstock:2014jp,Williams:2014ud}.

We took advantage of recent fires in alpine open heathland, a plant community that occupies ca. 25\% of the Australian alpine landscape \citep{Williams:2014ud}. Under global warming, this plant community is likely to encroach upon grasslands, wetlands and herbfields \citep[including the nationally endangered snow patch herbfields;][]{Williams:2015gj}, and is itself susceptible to shrub thickening \citep{McDougall:2003gb}. To determine drivers of shrub establishment and how warmer post-fire conditions affect shrub vital rates, we combined field observations of post-fire shrub seedling density with a warming experiment that used seedlings of four dominant Australian evergreen fire-killed shrubs: \textit{Grevillea australis} (Proteaceae; a tall shrub), \textit{Asterolasia trymalioides} (Rutaceae; a prostrate shrub),  \textit{Phebalium squamulosum} (Rutaceae; a tall shrub) and \textit{Prostanthera cuneata} (Lamiaceae; a tall shrub).

This combination of data sources allowed us to quantify some of the outstanding unknowns in the hypothesised climate-fire-shrub feedback loop (Fig. \ref{fig1}a). Specifically, we examined: 
\begin{enumerate}\itemsep 0pt
  \item Landscape scale drivers of shrub seedling establishment.
  \item How warmer post-fire conditions influence rates of grass recovery and shrub seedling recruitment, growth and mortality.
  \item How proximity to grass affects shrub seedling demographic rates; and whether such grass effects change under warmer post-fire conditions.
\end{enumerate}

@@

A summary of study sites, experimental design and approach to data analysis is given below. Full details are presented in the Supporting Information.

@@

The Australian Alps are an ideal place to examine potential fire-climate-vegetation feedbacks in high mountain environments because Australian alpine ecosystems have all the elements needed to demonstrate such potential feedbacks. Here, the Alps are subject to recurrent landscape fire, and were burnt extensively in 2003 \citep{Williams:2006tl,Williams:2014ud}. The vegetation consists of a range of life forms (graminoids, other herbs, shrubs), and variation in life-form dominance between and within vegetation types is a strong determinant of landscape flammability \citep{Williams:2006tl,Fraser:NHT5Kx_W}. There is also evidence that both fire regimes and the relative abundance of lifeforms in the Australian Alps have changed as a consequence of recent climate change \citep{Bradstock:2014jp,Wahren:2013ji,Camac:2015hja}.

We investigated shrub-grass dynamics in open heathland vegetation on the Bogong High Plains, Australia. Open heathland is a common Australian alpine vegetation type, occupying approximately 25\% of the treeless landscapes above 1600m in the Australian Alps. It is dominated by the shrub \textit{Grevillea australis} (shrub cover generally 20\textendash 60\%) with inter-shrub regions occupied by snow grasses (\textit{Poa hiemata} and \textit{Poa costiniana}) and other herbaceous species. It is an ecotone between closed heathland (\textgreater 70\% shrub cover) and tussock grassland (shrub cover < 20\%). \textit{Grevillea australis} is an obligate seeding shrub, and establishment of seedlings is dependent on disturbance creating bare ground in the grass\textendash herb sward \citep{Williams:1987ws}. Relative to tussock grassland, open heathland is a highly flammable plant community \citep{Williams:2006tl,Fraser:NHT5Kx_W}. A consequence of this differential flammability lead to the 2003 wildfires on the Bogong High Plains burning approximately 60\% of open heathland and only 13\% of tussock grasslands. Following these fires, seedling regeneration of \textit{Grevillea australis} and other shrubs was prolific \citep{Williams:2014ud}.

@@

We used 40 open heathland monitoring sites established after the 2003 fires \citep{Williams:2006tl,Camac:2013eb}. These sites consisted of 30 burnt sites and 10 sites known to be unburnt for over 70 years. The disparity in sampling effort between burnt and unburnt was due to very few open heathland sites of sufficient size (0.25 ha) remaining unburnt after both the 2003 and 2006 fires. In the summer of 2011-12, at each site, seedling density/m$^2$ was estimated using 40 quadrats, each 1 m$^2$, that were evenly distributed in groups of 10 along four 50 m transects, with 10 m between transect lines, subsampling an area of 2000 m$^2$. Within plots we recorded the number and maximum height of \textit{Grevillea} and \textit{Asterolasia} seedlings. For unburnt sites we counted the number of mature \textit{Grevillea} plants (\textgreater 0.5 m$^2$) within 5 m of each transect. In burnt sites, this required counting the number of skeletons (there were no living adults at any burnt site) that still persisted post-fire. We were unable to estimate numbers of adult \textit{Asterolasia} because this species does not have a persistent woody skeleton post-fire. Site level data, elevation and Topographic Wetness Index \citep[TWI; a measure of plant available water,][]{Moore:1993fk} were obtained from a 30 m resolution digital elevation model. Lastly, for burnt sites, fire severity was estimated by twig diameters \citep{Whight:1999vw}, collected immediately after the 2003 fires \citep{Williams:2006tl}.

@@

Climate change is expected to alter biotic interactions because their strength and direction depend strongly on climatic conditions, particularly in alpine and arctic ecosystems \citep{Callaway:2002vb, Klanderud:2005uw}. Here, we assess the interactive effects of warming and grass proximity on the growth and survival of \textit{Grevillea} seedlings transplanted into various sized inter-tussock gaps. We detected a strong positive effect of warming treatment on growth rates and a marginal, yet not significant, decrease in mortality (Fig. \ref{fig5}). However, we did not detect significant inter-tussock gap size effects or an interaction between gap size and warming treatment on either growth or mortality rates (i.e. coefficient credible intervals overlap zero).

@@

Despite having little impact on shrub seedling growth and mortality rates, tussock grass may still limit shrub recruitment, and thus, shrub expansion, by infilling post-fire bare ground gaps (whether by vegetative growth or seedlings) faster under warmer conditions. Using six years of post-fire inter-tussock gap size changes in warmed and unwarmed plots we found that gaps were being infilled by tussock grasses (Fig. \ref{fig6}). However, the rate at which this occurred was very slow, with a 10 cm radius gap predicted to decrease by approximately \Sexpr{gap_size(gap_dynamic_model, 'otc')} cm over a ten year period. We also detected no significant effect of a \Sexpr{micro_clim(overall_microclimate_summary,'CAmbient_Temp','mean','grand_mean')}\textdegree C temperature rise on the rate of infilling.

@@

In this study we have quantified several unknown interactions between fire, shrub-grass relationships and climate to extend a conceptual model of alpine shrub dynamics under climate change (Fig. \ref{fig1}b). In particular, we have shown how these interactions strengthen a hypothesized feedback loop that can rapidly increase shrub cover in the Australian Alps, and potentially other alpine and tundra ecosystems \citep{Higuera:2008dw}. While high cover of tussock grass is known to inhibit the establishment of shrub seedlings \citep{Williams:1987ws}, more frequent and severe fire caused by changing climate, will create bare ground required for shrub seedlings to establish. Our results, highlight that given a shrub seedling establishes in bare ground, its growth and survival rates are not affected by proximity to tussock grass, irrespective or warming. More importantly, however, our results indicate that for tall shrubs, a 1\textdegree C increase in mean growing season temperature resulted in a doubling of growth rates and a potential increase in survival. These demographic effects will ultimately manifest in shrub thickening within and at shrub-grassland/herbfield boundaries. It will also facilitate shrub establishment into non-shrubby communities by increasing seed sources at the boundary (i.e. more mature plants) and decreasing dispersal distances.

Evidence from field studies of burning patterns \citep{Williams:2006tl}, quantification of fuel mass and architecture \citep{Fraser:NHT5Kx_W} and paleoecological studies \citep{Higuera:2008dw} indicate that shrubs are the most flammable component of alpine and tundra ecosystems \citep[particularly moreso then tussock grassland; ][]{Fraser:NHT5Kx_W}. Consequently, our results indicate that flammable fuel loads will accumulate twice as fast under a warmer environment. This is likely to further strengthen the feedback \citep{Higuera:2008dw, Higuera:2009ga} by increasing the frequency and severity of fires, which then creates more shrub recruitment opportunities both within and beyond shrub-grassland boundaries that may persist for decades.

While we have addressed several unknowns associated with this feedback between climatic warming, shrubs and fire, there are others we have not addressed that may also strengthen, weaken or break this feedback. The most obvious mechanism that will break this cycle involves shorter fire intervals that prevent fire-killed shrubs reaching reproductive age and thereby exhausting the seedbank \citep{Enright:2015em}. However, this scenario is unlikely for the majority of alpine (or tundra) landscapes, including those in Australia, where current fire intervals of 50 to 100 years would need to decrease to less than 20 years\textemdash the time estimated for the species in this study to reach reproductive maturity \citep{Williams:2008ba}. Furthermore, if, as hypothesised, reproductive output is related to plant size \citep{Wenk:2015ko}, then climatic warming may allow fire-killed species to reach reproductive maturity sooner, and consequently, may increase species resilience to short fire intervals. Nevertheless, these factors and others such as changes in snow pack \citep{Wipf:2009jh}, soil moisture \citep{Engler:2011vq} or herbivory \citep{Post:2008wz} require further research because they are likely to be altered by the interactive effects of climate and disturbance in unpredictable ways \citep{MyersSmith:2011fp}.

>>>>

\begin{abstract}
Climate change is expected to increase fire activity and woody plant encroachment in arctic and alpine landscapes. However, the extent to which these increases interact to affect changes in structure, function and composition of alpine ecosystems is unknown. Here we use field surveys and experimental manipulations to examine warming and fire effects on recruitment, growth and survival of seedlings of four dominant, evergreen fire-killed alpine shrubs. We found that fire substantially increased shrub seedling establishment (up to \Sexpr{density_ratios(greaus_seedling_density_model, 'Grevillea')}\textendash fold) and that experimental warming doubled tall shrub seedling growth rates and could potentially increase survival. Warming had no effect on shrub recruitment, post-fire infilling of gaps by a dominant alpine grass, or competitive effects of grass on shrub seedling growth and survival. These findings indicate that warming, coupled with more frequent or severe fires, will likely result in an increase in the cover and abundance of evergreen shrubs by increasing recruitment, doubling growth rates, and potentially increasing shrub survival. Thus, warming is likely to strengthen an existing feedback between woody species abundance and fire activity in alpine and tundra environments.
\end{abstract}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\section*{Introduction}
Accurately forecasting the effects of climatic warming on vegetation dynamics requires an understanding of the mechanisms by which climate and vegetation interact. Most forecasting models include the direct effects of climatic conditions on species distributions, but largely ignore disturbances, particularly their type, frequency and severity \citep{Guisan:2005gz,Tucker:2012dr}. Recurrent disturbance is integral to all ecosystems, strongly influencing fundamental demographic processes such as recruitment and mortality, and thus the composition and structure of plant communities and biomes worldwide \citep{Bond:2005eg, Turner:2010wo}. Disturbance regimes are changing as a consequence of climate change \citep{Turner:2010wo, Westerling:2011jp,Bradstock:2014jp}. Moreover, the effects of disturbance will interact with climate change, and thus may amplify or diminish how changes to temperature and moisture affect vegetation \citep{Post:2008wz,Camac:2015hja,Enright:2015em}

Feedbacks between climate change, vegetation and fire will be particularly important determinants of the trajectory of ecosystem change in coming decades \citep{Bowman:2009kp}. For example, in addition to the direct effects of warming on fire weather, changing climate may also alter fire regimes by altering vegetation productivity, structure and plant trait composition \citep{Keeley:2012wo,Matthews:2012gia,Bowman:2014dr}. Feedbacks, which may be positive or negative, have been documented in tropical savanna \citep{Hoffmann:2003br}, boreal forests \citep{Krawchuk:2011wr}, Mediterranean ecosystems \citep{Keeley:2012wo}, grasslands \citep{Flannigan:2009kl} and alpine and arctic environments \citep{Goetz:2007gj,Wookey:2009wo}.

@@

However, most studies in arctic and alpine areas have focused on plant responses in mature, undisturbed vegetation \citep{Briceno:2015iv}. Few climate-change field experiments have included disturbance as a factor \citep[but see:][]{Munier:2010cg,Graae:2010ir,Camac:2015hja}, or examined the influence of climate change on seedlings in post-disturbance environments. Although alpine and tundra vegetation has a strong capacity to regenerate vegetatively \citep{Billings:1968bn} there is mounting evidence that seedling regeneration is important in these ecosystems \citep{Venn:2009bw, Briceno:2015iv}, particularly for woody species \citep{Camac:2013eb,Williams:2014ud}. Despite seedlings being highly vulnerable to climate and disturbance, they are the life stage that determines the long-term persistence of a species and its capacity to establish in new areas \citep{Walck:2011dm}. Little is known about how climate change affects seedling demography \citep{Briceno:2015iv}, particularly in alpine and tundra post-disturbance conditions. This is unfortunate, because if we wish to accurately predict future trajectories of vegetation change in these ecosystems, it is imperative we understand how seedlings respond to both changing climate and disturbance regimes \citep{Walck:2011dm, Briceno:2015iv}.

\section*{Conceptual feedback model}

In alpine and arctic ecosystems, warming experiments and long-term monitoring have documented significant increases in the growth and cover of woody species \citep{Sturm:2001vd,Walker:2006ht,MyersSmith:2011fp}. The frequency and extent of wildfires have also increased in these environments over recent decades, a trend expected to continue \citep{Westerling:2006ve, Flannigan:2009kl, Qiu:2009th, Bradstock:2014jp}. In alpine and tundra ecosystems, current evidence from long-term and experimental studies indicate that shrub recruitment and encroachment is highest in disturbed areas \citep{Williams:1987ws, Batllori:2009fe, Frost:2013tm} and lowest in areas with minimal bare ground cover, such as undisturbed alpine tussock grassland \citep{Williams:1987ws}. Current evidence indicates that climatic warming is likely to increase the growth rates of woody species \citep{Arft:1999iz, Elmendorf:2012ba, MyersSmith:2011fp} and that shrubs are the most flammable vegetation component of these ecosystems \citep{Williams:2006tl, Higuera:2009ga, Fraser:NHT5Kx_W}. Thus, increases in the cover of woody species in alpine and arctic environments is likely to increase the flammability of these ecosystems.

The results of these studies suggest a positive feedback between warmer temperatures, woody species and fire in alpine and tundra environments (Fig. \ref{fig1}a). Warmer temperatures are likely to lead to more frequent or severe fire. This is likely to increase recruitment opportunities for woody species and increase shrub thickening both within and beyond shrub boundaries \citep{Racine:2004tn,Lantz:2013fh}. If this effect is coupled with an increase in the growth and mortality of shrub seedlings, highly flammable fuels will accumulate faster, have higher landscape connectivity, and ultimately lead to increases in the likelihood of fire \citep{Higuera:2008dw, Higuera:2009ga,MyersSmith:2011fp}. Thus, warming could strengthen an existing climate-disturbance feedback, that not only has the potential to cause rapid changes in the composition and structure of alpine and arctic vegetation, but also has serious social, biodiversity and carbon sequestration consequences \citep{Mack:2011ew}.

\begin{figure}[!h]
\includegraphics{figures/fig1.pdf}
\caption{Conceptual diagram illustrating positive warming-shrub-fire feedback. (a) Hypothesised positive feedback loop between fire, climatic warming and landscape flammability. Solid lines = known mechansims; Dashed lines = mechansims that may exacerbate or diminish this feedback but which we have a paucity of information on. (b) Feedback updated based on new evidence obtained from experimental and field surveys conducted in this study. Also includes an example of a potential, but unlikely, effect (red dashed line) that could break this feedback. Blue = positive relationships, Red = negative relationships.}
\label{fig1}
\end{figure}

While paleoecological studies have indicated that such a feedback may exist in the arctic \citep{Higuera:2008dw, Higuera:2009ga}, there is a paucity of information on what may strengthen, mitigate or break this feedback. For example, we have little information on whether fire will stimulate shrub seedling recruitment in alpine or tundra ecosystems, nor do we know how rates of seedling growth and survival will be affected under warmer, more exposed, post-fire conditions. Furthermore, we do not know how tussock grasslands will respond to warmer post-fire conditions, how grasses affect vital rates of shrub seedlings, or whether such effects are altered by warmer post-fire conditions.

Here, we examine these unknowns and their impacts on this proposed feedback between climatic warming, shrubs and fire in Australian alpine vegetation. As with other tundra ecosystems, Australian alpine landscapes consist of a mosaic of vegetation types dominated by either herbaceous or woody species \citep{Williams:2014ud}. They have also experienced significant changes in climate. Since 1979, mean growing season temperatures have risen by approximately 0.4\textdegree C and annual precipitation has fallen by 6\% \citep{Wahren:2013ji}, with a consequent decline in snow pack depth \citep{SanchezBayo:2013ci}. These climatic changes have been correlated with a 10 to 20\% increase in shrub cover and a 25\% decline in graminoids cover \citep{Wahren:2013ji}. Much of the Australian Alps has also been burnt by recent (2003 and 2006) wildfires, the frequency and/or severity of which are expected to increase as a consequence of further climate change \citep{Bradstock:2014jp,Williams:2014ud}.

We took advantage of recent fires in alpine open heathland, a plant community that occupies ca. 25\% of the Australian alpine landscape \citep{Williams:2014ud}. Under global warming, this plant community is likely to encroach upon grasslands, wetlands and herbfields \citep[including the nationally endangered snow patch herbfields;][]{Williams:2015gj}, and is itself susceptible to shrub thickening \citep{McDougall:2003gb}. To identify some of the biotic and abiotic factors that affect shrub establishment and how warmer post-fire conditions affect shrub seedling vital rates, we combined field observations of post-fire seedling densities with a warming experiment that used seedlings of four dominant Australian evergreen fire-killed shrubs: \textit{Grevillea australis} (Proteaceae; a tall shrub), \textit{Asterolasia trymalioides} (Rutaceae; a prostrate shrub), \textit{Phebalium squamulosum} (Rutaceae; a tall shrub) and \textit{Prostanthera cuneata} (Lamiaceae; a tall shrub).

This combination of data allowed us to quantify the following unknowns in the hypothesised climate-fire-shrub feedback loop (Fig. \ref{fig1}a): 
\begin{enumerate}\itemsep 0pt
  \item Does fire stimulate shrub recruitment?
  \item Does warming alter the recruitment of shrubs post-fire?
  \item Does warming alter the growth and survival of shrub seedlings?
  \item Does proximity to tussock grass affect the growth and survival of shrub seedlings?
  \item Does warming alter the post-fire regeneration of tussock grass?
\end{enumerate}

@@

Below, we provide a summary of the experimental design and analysis. Full details are presented in the Supporting Information.

@@

The Australian Alps are an ideal place to examine potential fire-climate-vegetation feedbacks in high mountain environments because they have all the elements needed to demonstrate such potential feedbacks. The Alps are subject to recurrent landscape fire, and were burnt extensively in 2003 \citep{Williams:2006tl,Williams:2014ud}. The vegetation consists of a range of life forms (graminoids, other herbs, shrubs), the abundance of which strongly determines landscape flammability \citep{Williams:2006tl,Fraser:NHT5Kx_W}. There is also evidence that both fire regimes and the relative abundance of life forms in the Australian Alps have changed as a consequence of recent climate change \citep{Bradstock:2014jp,Wahren:2013ji,Camac:2015hja}.

We investigated shrub-grass dynamics in open heathland vegetation on the Bogong High Plains, Australia. Open heathland is a common Australian alpine vegetation type, occupying approximately 25\% of the treeless landscapes above 1600m in the Australian Alps. It is dominated by the shrub \textit{Grevillea australis} (shrub cover generally 20\textendash 60\%) with inter-shrub regions occupied by snow grasses (\textit{Poa hiemata} and \textit{Poa costiniana}) and other herbaceous species. It is an ecotone between closed heathland (\textgreater 70\% shrub cover) and tussock grassland (shrub cover < 20\%). \textit{Grevillea australis} is an obligate seeding shrub, and establishment of seedlings is dependent on disturbance creating bare ground in the grass\textendash herb sward \citep{Williams:1987ws}. Relative to tussock grassland, open heathland is a highly flammable plant community \citep{Williams:2006tl,Fraser:NHT5Kx_W}. A consequence of this differential flammability led to the 2003 wildfires on the Bogong High Plains burning approximately 60\% of open heathland and only 13\% of tussock grasslands. Following these fires, seedling regeneration of \textit{Grevillea australis} and other shrubs was prolific \citep{Williams:2014ud}.

@@

We used 40 open heathland monitoring sites established after the 2003 fires \citep{Williams:2006tl,Camac:2013eb}. These sites consisted of 30 burnt sites and 10 sites known to be unburnt for over 70 years. The disparity in sample number between burnt and unburnt treatments was due to very few open heathland sites of sufficient size (0.25 ha) remaining unburnt after both the 2003 and 2006 fires. In the summer of 2011-12, at each site, seedling density/m$^2$ was estimated using 40 quadrats, each 1 m$^2$, that were evenly distributed in groups of 10 along four 50 m transects, with 10 m between transect lines, subsampling an area of 2000 m$^2$. Within plots we recorded the number and maximum height of \textit{Grevillea} and \textit{Asterolasia} seedlings. For unburnt sites we counted the number of mature \textit{Grevillea} plants (\textgreater 0.5 m$^2$) within 5 m of each transect. In burnt sites, this required counting the number of skeletons (there were no living adults at any burnt site) that still persisted post-fire. We were unable to estimate numbers of adult \textit{Asterolasia} because this species does not have a persistent woody skeleton post-fire. Site level data, elevation and Topographic Wetness Index \citep[TWI; a measure of plant available water,][]{Moore:1993fk} were obtained from a 30 m resolution digital elevation model. Lastly, for burnt sites, fire severity was estimated by twig diameters \citep{Whight:1999vw}, collected immediately after the 2003 fires \citep{Williams:2006tl}.

@@

Climate change is expected to alter biotic interactions because their strength and direction depend strongly on climatic conditions, particularly in alpine and arctic ecosystems \citep{Callaway:2002vb, Klanderud:2005uw}. Here, we assess the interactive effects of warming and grass proximity on the growth and survival of \textit{Grevillea} seedlings transplanted into various sized inter-tussock gaps. We detected a strong positive effect of warming treatment on growth rates and a marginally non-significant decrease in mortality (Fig. \ref{fig5}). However, we did not detect significant inter-tussock gap size effects or an interaction between gap size and warming treatment on either growth or mortality rates (i.e. coefficient credible intervals overlap zero).

@@

Despite having little impact on shrub seedling growth and mortality rates, tussock grass may still limit shrub recruitment, and thus, shrub expansion, by infilling post-fire bare ground gaps (whether by vegetative growth or seedlings) faster under warmer conditions. Using six years of post-fire inter-tussock gap size changes in warmed and unwarmed plots, we found that gaps were being infilled by tussock grasses (Fig. \ref{fig6}). However, the rate at which this occurred was very slow, with a 10 cm radius gap predicted to decrease by approximately \Sexpr{gap_size(gap_dynamic_model, 'otc')} cm over a ten year period. We also detected no significant effect of a \Sexpr{micro_clim(overall_microclimate_summary,'CAmbient_Temp','mean','grand_mean')}\textdegree C temperature rise on the rate of infilling.

@@

In this study we have quantified several unknown interactions between fire, shrub-grass relationships and climate to extend a conceptual model of alpine shrub dynamics under climate change (Fig. \ref{fig1}b). In particular, we have shown how these interactions strengthen a hypothesized feedback loop that can rapidly increase shrub cover in the Australian Alps, and potentially other alpine and tundra ecosystems \citep{Higuera:2008dw}. While high cover of tussock grass is known to inhibit the establishment of shrub seedlings \citep{Williams:1987ws}, more frequent and severe fire caused by changing climate, will create bare ground required for shrub seedlings to establish. Our results highlight that if a shrub seedling establishes in bare ground, its growth and survival rates are not affected by proximity to tussock grass, irrespective or warming. More importantly, our results indicate that for tall shrubs, a 1\textdegree C increase in mean growing season temperature resulted in a doubling of growth rates and a potential increase in survival. These demographic effects will manifest in shrub thickening within and at shrub community boundaries, and ultimately, facilitate shrub establishment into non-shrubby communities.

Evidence from field studies of burning patterns \citep{Williams:2006tl}, quantification of fuel mass and architecture \citep{Fraser:NHT5Kx_W} and paleoecological studies \citep{Higuera:2008dw} indicate that shrubs are the most flammable component of alpine and tundra ecosystems \citep[significantly more than tussock grassland; ][]{Fraser:NHT5Kx_W}. Consequently, our results indicate that flammable fuel loads will accumulate twice as fast under a warmer climate. This is likely to further strengthen the feedback \citep{Higuera:2008dw, Higuera:2009ga} by increasing the frequency and severity of fires, which then creates more shrub recruitment opportunities both within and beyond shrub-grassland boundaries. Such changes may persist for decades.

While we have addressed several unknowns associated with this feedback between climatic warming, shrubs and fire, there are others we have not addressed that may also strengthen, weaken or break this feedback. The most obvious mechanism that will break this cycle involves shorter fire intervals that prevent fire-killed shrubs reaching reproductive age and thereby exhausting the seedbank \citep{Enright:2015em}. However, this scenario is unlikely for the majority of alpine (or tundra) landscapes, including those in Australia, where current fire intervals of 50 to 100 years would need to decrease to less than 20 years\textemdash the time estimated for the species in this study to reach reproductive maturity \citep{Williams:2008ba}. Furthermore, if, as hypothesised, reproductive output is related to plant size \citep{Wenk:2015ko}, then climatic warming may allow fire-killed species to reach reproductive maturity sooner, and consequently, may increase species resilience to short fire intervals. Nevertheless, these factors and others such as changes in snow pack \citep{Wipf:2009jh}, soil moisture \citep{Engler:2011vq}, herbivory \citep{Post:2008wz} and adaptation \citep{Byars:2009fd, Hoffmann:2011fp} require further research because they are likely to be altered by the interactive effects of climate and disturbance in unpredictable ways \citep{MyersSmith:2011fp}.

By focusing on seedlings, the life stage most vulnerable to climate and disturbance, and which determines the long-term persistence of shrub species and its capacity to establish in new areas \citep{Walck:2011dm, Briceno:2015iv}, our findings provide a possible explanation as to why shrub cover has increased in the Australian Alps, often at the expense of grasslands \citep{McDougall:2003gb,Wahren:2013ji}. An explanation that may also apply to other arctic and alpine ecosystems \citep{Racine:2004tn,MyersSmith:2011fp,Frost:2013tm,Lantz:2013fh}. But more importantly, our results provide evidence for the underlying processes that could result in a warming-fire-shrub feedback that has been hypothesized in paleoecological studies \citep{Higuera:2008dw, Higuera:2009ga}. Based on current observations and predictions, average global temperature has already increased by 0.85\textdegree C since 1880 and is expected to rise by as much as 4.8\textdegree C by 2100 \citep{IPCC:2013vv}. In alpine and tundra environments, temperatures \citep{ChapinIII:2005js}, shrub cover \citep{MyersSmith:2011fp} and the frequency and severity of fire \citep{Westerling:2006ve, Flannigan:2009kl,Qiu:2009th, Bradstock:2014jp} have all increased across the globe in the last few decades as a consequence of rising temperatures. These changes mean that the warming-shrub-fire feedback loop identified here is likely to have already strengthened. Thus, other non-woody communities will become shrubbier and more prone to fire, the effects of which, will have significant consequences for carbon sequestration, water supply and biodiversity.
